### 2017-11-15 submission/manuscript.Rmd
change title
finish Will's suggestions for discussion and conclusion

# The Microbiota and Individual Community Members in Colorectal Cancer: Is There a Common Theme?

@@

Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma. In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma. The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle. Although there may be localized changes that do depend on the driver-passenger model, supported by an increased relative risk for one, two, and four positive CRC associated genera in tissue [Figure 3C], there may be other processes involved that ultimately exacerbates the condition from a subtle localized change to a global community one. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.

Although there are still questions that need to be answered for the microbiota and carcinoma, a clearer framework is beginning to develop as to how this occurs. The role of the microbiota in adenoma is still not clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here. It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control. Most of the studies analyzed have sufficient power to detect these type of changes. In contrast, our data suggests that the adenoma effect size is relatively small. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences consistently occur in adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma formation.

@@

By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the importance of the bacterial community in both adenoma and carcinoma. We are also able to provide support for the driver-passenger model in the context of carcinoma. However, within the context of adenoma it is less clear that this relationship exists. These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in the community.

>>>>

# Investigating the Microbiota and Colorectal Cancer: The Importance of Community

@@

With regards to carcinoma, our results support the driver-passenger model within the framework of the transition from adenoma to carcinoma. Conversely, using the present data, we can only theorize observations related to adenoma development fit this model. The changes that occur at this timepoint are small and possibly focal to the adenoma itself. The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly. Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change to a change in the global community. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. It is possible to hypothesize that at early stages of the diease, the host interacts with these subtle changes and catalyzes what ultimately leads to a thoroughly dysfunctional community.

Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs. From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 6]. Conversely, the role of the microbiota in adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls. This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma formation.

@@

By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in both adenoma and carcinoma. We are also able to provide support for the driver-passenger model in the context of carcinoma. However, within the context of adenoma, it is less clear that this relationship exists. These observations highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community. Although there are power limitations associated with adenoma, this report highlights the strong influence the microbiota has on CRC development.

### 2017-11-15 submission/manuscript.tex
finish Will's suggestions for discussion and conclusion

Our carcinoma observations fit the driver-passenger model and support
this concept within the framework of the transition from adenoma to
carcinoma. In contrast, with the present data we can only suggest that
the adenoma observations might fit with this model but the changes that
occur at this timepoint are small and possibly focal to the adenoma. The
stepwise decrease in diversity suggests that the adenoma community is
not normal but this change is subtle. Although there may be localized
changes that do depend on the driver-passenger model, supported by an
increased relative risk for one, two, and four positive CRC associated
genera in tissue {[}Figure 3C{]}, there may be other processes involved
that ultimately exacerbates the condition from a subtle localized change
to a global community one. The poor performance of the Random Forest
models for classifying adenoma based only on the microbiota would
suggest that this is the case. It is possible to hypothesize that at
early stages of the diease, how the host interacts to these subtle
changes could be the catlyst that ultimately leads to this larger global
dysfunctional community.

Although there are still questions that need to be answered for the
microbiota and carcinoma, a clearer framework is beginning to develop as
to how this occurs. The role of the microbiota in adenoma is still not
clear and part of the reason this may be is because many studies are not
powered effectively to observe the small changes reported here. It is
realistic to suspect that many changes in carcinoma could easily result
in effect sizes that are 30\% or more between the case and control. Most
of the studies analyzed have sufficient power to detect these type of
changes. In contrast, our data suggests that the adenoma effect size is
relatively small. None of the studies analyzed were properly powered to
detect a 10\% or lower change between case and controls and this may
well be the range in which differences consistently occur in adenoma.
Future studies investigating adenoma and the microbiota need to take
these factors into consideration if we are to work out the role of the
microbiota in adenoma formation.

>>>>

With regards to carcinoma, our results support the driver-passenger
model within the framework of the transition from adenoma to carcinoma.
Conversely, using the present data, we can only theorize observations
related to adenoma development fit this model. The changes that occur at
this timepoint are small and possibly focal to the adenoma itself. The
stepwise decrease in diversity suggests that the adenoma community is
not normal but has changed subtly. Although there appears to be
localized changes that do depend on the driver-passenger model, as
supported by an increased RR for one, two, and four positive
CRC-associated genera in tissue {[}Figure 3C{]}, there may be other
processes at play that ultimately exacerbate the condition from a subtle
localized change to a change in the global community. The poor
performance of the Random Forest models for classifying adenoma based
only on the microbiota would suggest that this is the case. It is
possible to hypothesize that at early stages of the diease, the host
interacts with these subtle changes and catalyzes what ultimately leads
to a thoroughly dysfunctional community.

Although there are still questions that need to be answered regarding
the microbiota and carcinoma, a clearer framework for their relationship
is beginning to develop as to how this occurs. From our observations
many changes in carcinoma could easily result in effect sizes that are
30\% or more between the case and control and most studies analyzed have
sufficient power to detect these types of changes {[}Figure 6{]}.
Conversely, the role of the microbiota in adenoma is less clear and part
of the reason this may be is because many studies are not powered
effectively to observe the small changes reported here. None of the
studies analyzed were properly powered to detect a 10\% or lower change
between case and controls. This small effect size range may well be the
scope in which differences consistently occur in adenoma due to the
subtle changes in community that occur between control and adenoma.
Future studies investigating adenoma and the microbiota need to take
these factors into consideration if we are to work out the role of the
microbiota in adenoma formation.
